Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-05-21 doc/paper.Rmd
edits up to Results

## Motivation
Same source analyses are a major part of a Forensic Toolmark Examiner's job. In current practice, examiners  make these comparisons by visual inspection under a comparison microscope and come to one of the following four conclusions: identification, inconclusive, elimination or unsuitable for examination~\citep{afte-toolmarks1998}. These conclusions are made on the basis of "unique surface contours" of the two toolmarks being in "sufficient agreement" \citep{afte-toolmarks1998}. AFTE describes the term "sufficient agreement" as the possibility of another tool producing the markings under comparison, as practically impossible \citep{afte-toolmarks1998}. Potential subject bias in the assessment as well as the lack of specified error rates are the main points of criticisms  first raised by the National Research Council in 2009 \citep{NAS:2009} and later emphasized further by the President's Council of Advisors on Science and Technology \citep{pcast2016}.

@@

Various toolmarks have been studied in the literature (see \autoref{tab:toolmarks-ER} for an overview):  \citet{manytoolmarks1} and \citet{chumbley} have been analyzing screwdriver marks  digitized using a profilometer;  \citet{manytoolmarks2} have investigated 3D marks from screwdriver, tongue and groove pliers captured using a confocal microscope; \citet{afte-chumbley} have been investigated digitized marks from slip-joint pliers generated by a surface profilometer.

@@

Striation marks on bullets are made by impurities in the barrel. As the bullet travels through the barrel, these imperfections leave "scratches" on the bullet surface. Typically, only striation marks in the land engraved areas (LEAs) are considered \citet{afte-article1992}. Bullet lands are depressed areas between the  grooves made by the rifling action of the barrel. Compared to toolmarks made by screwdrivers striation marks on bullets are typically much smaller, both in length and in width. Bullets also have a curved cross-sectional topography. Figure \ref{fig:rgl} shows us how the signature from a bullet land (bottom) lines up with the image of the land (top) from which it was extracted. We can also see in the figure how the depth and relative position of the striation markings seen in the image are interpreted as the signature.

Bullet matching methods are usually based on these associated signatures. \citet{chu2013} use an automatic method for counting consecutive matching striae (CMS). The authors report an error rate of 52% of the known same source lands comparisons as misidentified (false negative) and zero false positives for known different source lands. \citet{ma2004} and \citet{vorburger2011} discuss CCF (cross-correlation function) and its discriminating power and applicability for same-source analyses of bullets, but do not provide any error rates in their discussion. \citet{aoas} use multiple features like CCF, CMS, D (distance measure) etc in a random forest based method and compare every land against every other land of  digitised versions of Hamby 252 and Hamby 44 \citep{hamby} published on the NIST Ballistics Database \citep{nist}. The authors report an out-of-bag overall error rate of 0.46%, comprised of an error rate of 30.05% of  same-source pairs that were not identified and an error rate of 0.026% of different-source pairs that were incorrectly identified as same-source.

The Chumbley score  provides us with another approach in the same-source assessment of bullet striation marks. \citet{chumbley} compare two toolmarks for same-source. The data for this study was obtained from 50 sequentially manufactured screwdriver tips. \citet{chumbley} report error rates for markings made by the tips at different angles. For markings made at 30 degree the authors  report an average false negative error rate of 0.089  and an average false positive error rate of 0.023. For other angles of 60 and 85 degrees the false negatives error rate is 0.09 while the rate of false positives decreases to 0.01. The paper by \citet{hadler} is based on the same data but the authors focus on  markings made under the same angle. The error rates associated with the deterministic version of the score are  0.06 for false  negatives and a false positive error rate of 0.

@@

Comparisons of striae from bullets are usually based on comparisons of striae in land engraved areas, which are extracted in form of cross sections, called *profiles* \citep{aoas,ma2004}. From profiles bullet *signatures* \citep{chu2013,aoas} are extracted as residuals of a loess fit or Gaussian filter. This effectively removes topographic structure from the data in the attempt to increase the signal to noise ratio. The span of the loess fit was found using cross-validation, as described by \citet{aoas}.

There are two sources of scans for sets from the Hamby study available to us: scans of Hamby 44 and Hamby 252 are available from the NIST database \citep{nist}. Hamby 44 has also been made available to us and has been scanned locally for CSAFE at the Roy J.\ Carver High Resolution Microscopy Facility using a Sensofar confocal light microscope. 
Scans in the NIST database are made with a NanoFocus at 20x magnification. The resolutions of the two instruments are different: the NIST scans are taken at a resolution of 1.5625 $\mu m$ per pixel, while the CSAFE scans are available at a resolution of  0.645 $\mu m$ per pixel.
The length of an average bullet land from Hamby (9 mm Ruger P85) is about 2 millimeter, resulting in signatures of about 1200 pixels for NIST scans, and about 3000 pixels for CSAFE scans.

@@

The Chumbley score algorithm takes input in form of two digitized toolmarks. The toolmark is in form of $z(t)$ which is a spatial process for location indexed by $t$. $t$ here denotes equally spaced pixel locations for the striation marks under consideration, $t = 1, ..., T$. Let further $z^s(t)$ denote a vector of markings of length $s$ starting in location  $t$.

Let $x(t_1)$, $t_1 = 1,2,...T_1$ and $y(t_2)$, $t_2 = 1,2...T_2$ be two digitized toolmarks (where $T_1$ and $T_2$ are not necessarily equal). 
<!--The vectorized processes representing two sets of striaes are therefore shown as $x(t_1)$, $t_1 = 1,2,...T_1$ and $y(t_2)$, $t_2 = 1,2...T_2$. Here $x$ and $y$ denote two instances of the process $z(*)$ which means two striation marks, while the indexing $t$ or pixel range is shown as $t_1$ and $t_2$ for $x$ and $y$ respectively. This representation of $t$ as $t_1$ and $t_2$ lets us identify them as either of same or different lengths.--> 
The toolmarks under consideration are potentially from two different sources or the same source. $T_1$ and $T_2$, as represented above, are the final pixel indexes of each marking and therefore give the respective lengths of the markings.

In a pre-processing step  the two markings are smoothed using a lowess \citep{lowess} with coarseness parameter $c$. 
 Originally, this smoothing is intended to remove drift and (sub)class characteristics from individual markings, however, in the setting of matching bullet striae, we can also make use of this mechanism to separate bullet curvature in profiles from signatures before matching signatures.

After removing sub-class structure, 
the Chumbley scores is calculated in two steps: an optimization step and a validation step.
In the optimization step, the two markings are aligned horizontally such that within a pre-defined window of length $w_o$ the correlation between $x(t_1)$ and $y(t_2)$ is maximized:
\[
\left(t_1^o, t_2^o\right) = \mathop{\arg \max}\limits_{1 \le t_1 \le T_1, 1 \le t_2 \le T_2} \text{cor} \left(x^{w_o} (t_1), y^{w_o}(t_2) \right)
\]
This results in an optimal vertical (in-phase) shift of $t_1^o - t_2^o$ for aligning the two markings. \hh{We will denote the relative optimal locations as $t_1^*$ and $t_2^*$, where $t_i^* = t_i^o/(T_i-w_o)$ for $i=1,2$, such that $t_1^*, t_2^* \in [0,1]$. Once (sub-)class characteristics are removed, the relative optimal locations should be distributed according to a uniform distribution in $[0,1]$. }

In the validation step,  two sets of windows of size $w_v$ are chosen from both markings (see Figure \ref{fig:win-comparison}). In the first set, pairs of windows are extracted from the two markings using the optimal vertical shift as determined in the first step, whereas for the second set the windows are extracted using a different (out-of-phase) shift.

@@

Same-shift pairs of length $w_v$ are defined in \citet{hadler} as all 
pairs $(s_i^{(1)}, s_i^{(2)})$ for all integer values $i$ for which both $s_i^{(1)}$ and  $s_i^{(2)}$ are defined. Similarly, different-shift pairs are defined as $(s_i^{(1)}, s_{-i-1}^{(2)})$ for all $i$ where both $s_i^{(1)}$ and  $s_{-i-1}^{(2)}$ are defined (see \autoref{sketch-same-diff}).

@@

```{r win-comparison, warning=FALSE, fig.width=8, fig.height=3.5, fig.cap="Two markings made by the same source. For convenience, the markings are moved into phase on the left and out-of phase on the right. In-phase (left) and out-of-phase (right) samples are shown by the light grey background. The Chumbley-score is based on a Mann-Whitney U test of the correlations derived from these two sets of samples."}
sigs_graphics <- readRDS("../data/sigs_generate_window.rds")
sigs_graphics$sigs_land_id_1_match <- sigs_graphics$sigs_land_id_1_match %>% group_by(land_id) %>% mutate(y = y-min(y, na.rm=T))

@@

For both same- and different-shift pairs correlations between the  markings are calculated. 
The intuition here is that for two markings from the same source the correlation for the in-phase sample should be high, while the correlations of the out-of-phase sample  provide a measure for the base-level correlation for non-matching marks of a given length $w_v$. The Chumbley score is then computed as a Mann Whitney U statistic to compare between in-phase sample and out-of-phase sample. 
In the original method proposed in \citet{chumbley} both in-phase and out-of-phase sample are extracted randomly, whereas \citet{hadler} proposed the above specified deterministic rules for both samples to make the resulting score deterministic while simultaneously avoiding overlaps within selected marks to ensure independence.

@@

Looking closer at \autoref{eq.start}, we see that by definition, some number of tests will fail to produce a result. \hh{The problem of failed tests is first mentioned in \citet{afte-chumbley}. Unfortunately, the authors do not provide any percentage of how many tests failed for their data.}
The algorithm fails to produce for two reasons: either the number of eligible same-shift pairs is zero, or the number of different-shift pairs is zero.

The number of same-shift pairs will be zero, if the optimal locations $t_1^{o}$ and $t_2^{o}$ are so far apart, that no segments of size $w_v$ are left on the same sides of the optimal locations, i.e. $t_1^{o} < w_v$ and $t_2^{o} > T_2-w_o-w_v$ or  $t_1^{o} < T_1- w_o - w_v$ and $t_2^{o} < w_v$, i.e. we have a failure rate of 
\[
P\left( t_1^{o} < w_v \ \cap \ t_2^{o} > T_2-w_o-w_v\right) + P\left( t_1^{o} < T_1- w_o - w_v \ \cap \ t_2^{o} < w_v\right).
\]
While we can assume that once (sub-)class characteristics are removed, optimal locations $t_i^{o}$ are uniformly distributed across the length of the profile, we cannot assume that $t_1^o$ and $t_2^o$ are independent of each other. In particular, for same-source profiles, we would expect a strong dependency between these locations, in which case a large difference between locations is unlikely. However, for different source matches, we can assume that  locations are independent.
In that case, we expect a test to fail with a probability of $\frac{2 w_v^2}{(T_1-w_o)(T_2-w_o)}$. For an average length of $T_i$ of 1200 pixels, $w_o = 120$ pixels and $w_v = 30$ pixels this probability is about 0.0015.

The number of possible different-shift pairs also depends on the location of  the  optimal locations $t_1^o$ and $t_2^o$. Whenever the optimal locations are close to the boundaries, the number of possible pairings decreases and reaches zero, if $t_i^{(o)} < w_v$ or  $t_i^{(o)} > T_i-w_o- w_v$. Assuming a correlation between optimal locations $t_1^o$ and $t_2^o$ of close to one for same-source profiles, this results in an expected rate of failure of $2 w_v / (T_i-w_o)$, or about 5.6\% for an average length of $T_i$ of 1200 pixels, $w_o = 120$ pixels and $w_v = 30$ pixels. Assuming independence in the optimal locations for different source profiles the expected probability for a failed test is, again, $\frac{2 w_v^2}{(T_1-w_o)(T_2-w_o)}$.

@@

We then define the pairs for different-shifts as
\begin{equation}\label{eq.diff2}
\left(s_{(j)}^{(1)}, s_{(I-j+1)}^{(1)} \right) \text{ for } j = 
\begin{cases}
1, ..., I & \text{ for even } I \\
1, ..., (I-1)/2, (I-1)/2 + 2, ..., I & \text{ for odd } I
\end{cases},
\end{equation}
i.e.\ for an odd number of same-shift correlations, we skip the middle pair for the different-shift correlations (see \autoref{sketch-diff-2}).
This pairing ensures that the number of different-shift pairings is the same or at most one less than the number of same-shift pairings in all tests.
\hh{In the remainder of the paper, we will refer to the algorithm defined by \citet{hadler} as {\bf (CS1)} and the suggested modified algorithm as {\bf (CS2)} and compare their performance on the available scans of the Hamby study.}
\begin{figure}[hbtp]
\centering
\includegraphics[width=.7\textwidth]{images/sketch-diff-2.png}
\caption{\label{sketch-diff-2}Sketch of adjusted different-shift pairings. At most one of the same-shift pairings can not be matched with a different-shift pair. }
\end{figure}

\begin{figure}[hbtp]
\centering
\includegraphics[width=0.62\textwidth]{images/same-shift-failure.png}

\caption{\label{same-shift-failure}Sketch of same-shift pairings  (top) when the lag is too large to accomodate a a vaildation window in either of the two signatures}
\end{figure}

\caption{\label{diff-shift-failure}Sketch of diff-shift pairings  (top) when the number of diff-shift computations is likely to 0}
\end{figure}

\begin{figure}[hbtp]
\centering
\includegraphics[width=.7\textwidth]{images/CS1-failure-CS2-success.png}
\caption{\label{CS1-failure-CS2-success}Sketch of a case where CS1 fails but CS2 does not fail}
\end{figure}

@@

```{r sigs-profiles, fig.width = 8, fig.height = 2.5, fig.cap="Bullet land profile (left) and the corresponding signature (right) for one of the lands of Hamby-44."}
p1 <- sigs_graphics[[1]] %>% filter(land_id==19) %>%
  ggplot(aes(x = y, y = value)) + geom_line() +
  theme_bw() +
  ylab(expression(paste("Depth (in ",mu,m,")", sep=""))) + 
  xlab(expression(paste("Relative location (in ",mu,m,")", sep=""))) + 
  ggtitle("Profile")

@@

## Setup
Both algorithms (CS1) and (CS2) are implemented in the R package `toolmaRk` \citep{toolmark}. We applied both methods to all pairwise land-to-land comparisons of the Hamby scans provided by NIST for a total of 85,491 land-to-land comparisons.

@@

Initially, the default settings suggested in \citet{hadler} were used: $w_o = 120$ pixels or about 190 $\mu m$ (ten percent of the average length of profiles) and coarseness $c = 0.25$. 
\autoref{fig:fails} shows the percentage of failed tests among the 85,491 land-to-land comparisons of the NIST data for different values of the validation window size $w_v$. For same-source lands up to 12.5 percent of the tests fail using CS1. The highest percentage of failed tests under CS2 is 1.3% for different source tests using a validation window size $w_v$ of 60 pixels.  \hh{Rates of expected failures are based on simulation runs using covariances between locations of same-source profiles of 0.854, and 0.120 for locations from different-source profiles, matching observed covariances for the Hamby scans. }\hh{ Observed failure rates are higher than expected. This might be due to remaining sub-class structure at a coarseness of 0.3 resulting in a distribution of optimal locations different from the assumed uniform. }

@@

\hh{\autoref{fig:profile-sketch} gives an overview of the effect of different coarseness parameters: from left to right, coarseness levels $c$ are varied in steps of 0.05 from 0.1 to 0.3. The top row shows  resulting signature after smoothing the profile shown in Figure~\ref{fig:sigs-profiles} with the coarseness specified. 
The histograms in the bottom row show the relative optimal location $t^*$. Optimal locations are distributed uniformly once (sub-)class characteristics are removed. However, for coarseness values of $c > 0.20$ we see quite distinct boundary effects: optimal locations $t^*$ are found at the very extreme ends of a profile more often than one would expect based on a uniform distribution.}

@@

<!-- The results suggest that the Nominal type I error $\alpha$ value shows dependence on the size of the window of optimization. For a given window of optimization the actual Type I error is comparable to the nominal level for only a select few validation window sizes and for comparable validation window sizes of 30 and 50 as done here, the actual type I error does not seem to vary as much as it varies with the optimization window sizes .
A Test Fail, i.e. tests in which a particular parameter setting did not return a valid result, happens, when the shift to align two signatures is so large, that the remaining overlap is too small to accommodate  windows for validation, depends on whether known-match or known non-matches has predictive value, with test results from different sources having a much higher chance to fail. On conducting an analysis of all known bullet lands using the adjusted chumbley algorithm, Type II error was identified to be least bad for window of validation 30 and window of optimization 120. In case of unsmoothed raw marks (profiles), Type II error increases with the amount of smoothing and least for LOWESS smoothing coarseness value about 0.25 or 0.3. In an effort to identify the level of adaptiveness of the algorithm, comparisons were made between signatures and profiles. Their comparison with respect to validation window size for a fixed optimization window size suggested that, profiles have a total error (i.e all incorrect classification of known-matches and known non-matches) greater than or equal to the total error of signatures for all sizes of validation window. Profiles also fail more number of times than signatures in a test fail (for different coarseness keeping windows fixed and also for different validation windows keeping coarseness fixed) which lets us conclude that the behaviour of the algorithm for the profiles instead of pre-processed signatures is not better. Finally it should be noted that the current version of the adjusted chumbley algorithm seems to fall short when compared to other machine-learning based methods \citet{aoas}, and some level of modification to the deterministic algorithm needs to be identified and tested that would reduce the number of incorrect classifications.

@@

>>>>

## Motivation
Same-source analyses are a major part of a Forensic Toolmark Examiner's job. In current practice, examiners  make these comparisons by visual inspection under a comparison microscope and come to one of the following four conclusions: identification, inconclusive, elimination or unsuitable for examination \citep{afte-toolmarks1998}. These conclusions are made on the basis of "unique surface contours" of the two toolmarks being in "sufficient agreement" \citep{afte-toolmarks1998}. AFTE describes the term "sufficient agreement" as the possibility of another tool producing the markings under comparison, as practically impossible \citep{afte-toolmarks1998}. Potential subject bias in the assessment as well as the lack of specified error rates are the main points of criticisms  first raised by the National Research Council in 2009 \citep{NAS:2009} and later emphasized further by the President's Council of Advisors on Science and Technology \citep{pcast2016}.

@@

Methods for matching marks for a variety of tools have been studied in the literature (see \autoref{tab:toolmarks-ER} for an overview):  \citet{manytoolmarks1} and \citet{chumbley} have been analyzing screwdriver marks  digitized using a profilometer;  \citet{manytoolmarks2} have investigated 3D marks from screwdriver, tongue and groove pliers captured using a confocal microscope; \citet{afte-chumbley} have been investigated digitized marks from slip-joint pliers generated by a surface profilometer.

@@

Striation marks on bullets are made by impurities in the barrel. As the bullet travels through the barrel, these imperfections leave "scratches" on the bullet surface (see top of \autoref{fig:rgl}). Typically, only striation marks in the land engraved areas (LEAs) are considered \citep{afte-article1992}. Bullet lands are depressed areas between the  grooves made by the rifling action of the barrel. Compared to toolmarks made by screwdrivers striation marks on bullets are typically much smaller, both in length and in width. Bullets also have a curved cross-sectional topography.

<!--\hh{signatures haven't been introduced yet}-->
<!--Figure \ref{fig:rgl} shows us how the signature from a bullet land (bottom) lines up with the image of the land (top) from which it was extracted. We can also see in the figure how the depth and relative position of the striation markings seen in the image are interpreted as the signature.-->

<!--Bullet matching methods are usually based on these associated signatures.-->
In same-source comparisons this curvature is usually removed using some form of Gaussian filter \citep{ma2004} or non-parametric smoothing \citep{aoas}. 
\citet{chu2013} use an automatic method for counting consecutive matching striae (CMS). The authors report an error rate of 52% for known same-source land comparisons to be (incorrectly) identified as different-source (false negative) and zero false positives for known different-source lands. \citet{ma2004} and \citet{vorburger2011} discuss CCF (cross-correlation function) and its discriminating power and applicability for same-source analyses of bullets, but do not provide any error rates in their discussion. \citet{aoas} use multiple features, such as CCF, CMS, D (distance measure), etc. in a random forest based method and compare every land against every other land of  digitised versions of Hamby 252 and Hamby 44 \citep{hamby} published on the NIST Ballistics Database \citep{nist}. The authors report an out-of-bag overall error rate of 0.46%, comprised of an error rate of 30.05% of  same-source pairs that were not identified and an error rate of 0.026% of different-source pairs that were incorrectly identified as same-source.

The Chumbley score  provides us with another approach in the same-source assessment of bullet striation marks. \citet{chumbley} compare two toolmarks for same-source. The data for this study was obtained from 50 sequentially manufactured screwdriver tips. \citet{chumbley} report error rates for markings made by the tips at different angles. For markings made at 30 degree the authors  report an average false negative error rate of 0.089  and an average false positive error rate of 0.023. For marks made under angles of 60 and 85 degrees, respectively, the false negatives error rate is 0.09 while the rate of false positives decreases to 0.01. The paper by \citet{hadler} is based on the same data but the authors focus on  markings made under the same angle. The error rates associated with the deterministic version of the score are reported as  0.06 for false  negatives and  0 for false positives.

@@

Comparisons of striae from bullets are usually based on comparisons of striae in land engraved areas, which are extracted in form of cross sections, called *profiles* \citep{aoas,ma2004}. From profiles bullet *signatures* \citep{chu2013,aoas} are extracted as residuals of a loess fit or Gaussian filter. This effectively removes topographic structure from the data in the attempt to increase the signal to noise ratio. \autoref{fig:rgl}  shows how the signature from a bullet land (bottom) lines up with the image of the land (top) from which it was extracted. We can see in the figure how the depth and relative position of the striation markings seen in the image are interpreted as peaks and valleys in the signature. <!--The span of the loess fit was found using cross-validation, as described by \citet{aoas}.-->

There are two sources of scans for sets from the Hamby study available to us: scans of Hamby 44 and Hamby 252 are available from the NIST database \citep{nist}. The physical Hamby 44 set has also been made available to us and has been scanned locally for CSAFE at the Roy J.\ Carver High Resolution Microscopy Facility using a Sensofar confocal light microscope. 
Scans in the NIST database are made with a NanoFocus at 20x magnification. The resolutions of the two instruments are different: the NIST scans are taken at a resolution of 1.5625 $\mu m$ per pixel, while the CSAFE scans are available at a resolution of  0.645 $\mu m$ per pixel.
The length of an average bullet land from Hamby (9 mm Ruger P85) is about 2 millimeter, resulting in signatures of about 1200 pixels for NIST scans, and about 3000 pixels for CSAFE scans.

@@

A digitized toolmark forms a spatial process $z(t)$ with location indexed by $t$. $t$ here denotes equally spaced pixel locations for the striation marks under consideration. For a toolmark consisting of $t$ pixels, $t = 1, ..., T$.
 Let further $z^s(t)$ denote a vector of markings of length $s$ starting in location  $t$.

The Chumbley score algorithm takes input in form of two digitized toolmarks:

Let $x(t_1)$, $t_1 = 1,2,...T_1$ and $y(t_2)$, $t_2 = 1,2...T_2$ be two digitized toolmarks (where $T_1$ and $T_2$, the lengths of the two marks, are not necessarily equal). 
<!--The vectorized processes representing two sets of striaes are therefore shown as $x(t_1)$, $t_1 = 1,2,...T_1$ and $y(t_2)$, $t_2 = 1,2...T_2$. Here $x$ and $y$ denote two instances of the process $z(*)$ which means two striation marks, while the indexing $t$ or pixel range is shown as $t_1$ and $t_2$ for $x$ and $y$ respectively. This representation of $t$ as $t_1$ and $t_2$ lets us identify them as either of same or different lengths.--> 
The toolmarks under consideration are potentially from two different-sources or the same-source. <!--$T_1$ and $T_2$, as represented above, are the final pixel indexes of each marking and therefore give the respective lengths of the markings. -->

```{r sigs-profiles, fig.width = 8, fig.height = 2.5, fig.cap="Bullet land profile (left) and the corresponding signature (right) for one of the lands of Hamby-44."}
sigs_graphics <- readRDS("../data/sigs_generate_window.rds")
sigs_graphics$sigs_land_id_1_match <- sigs_graphics$sigs_land_id_1_match %>% group_by(land_id) %>% mutate(y = y-min(y, na.rm=T))

p1 <- sigs_graphics[[1]] %>% filter(land_id==19) %>%
  ggplot(aes(x = y, y = value)) + geom_line() +
  theme_bw() +
  ylab(expression(paste("Depth (in ",mu,m,")", sep=""))) + 
  xlab(expression(paste("Relative location (in ",mu,m,")", sep=""))) + 
  ggtitle("Profile")

In a pre-processing step  the two markings are smoothed using a lowess \citep{lowess} with coarseness parameter $c$. 
 Originally, this smoothing is intended to remove drift and (sub)class characteristics from individual markings, however, in the setting of matching bullet striae, we can also make use of this mechanism to separate bullet curvature in profiles from signatures before matching signatures.  \autoref{fig:sigs-profiles} shows an example of a bullet land profile (left) and the corresponding signature (right).

After removing sub-class structure, 
the Chumbley score is calculated in two steps: an optimization step and a validation step.
In the optimization step, the two markings are aligned horizontally such that within a pre-defined window of length $w_o$ the correlation between $x(t_1)$ and $y(t_2)$ is maximized:
\[
\left(t_1^o, t_2^o\right) = \mathop{\arg \max}\limits_{1 \le t_1 \le T_1-w_o, 1 \le t_2 \le T_2-w_o} \text{cor} \left(x^{w_o} (t_1), y^{w_o}(t_2) \right)
\]
This results in an optimal vertical (in-phase) shift of $t_1^o - t_2^o$ for aligning the two markings. We will denote the relative optimal locations as $t_1^*$ and $t_2^*$, where $t_i^* = t_i^o/(T_i-w_o)$ for $i=1,2$, such that $t_1^*, t_2^* \in [0,1]$. Once (sub-)class characteristics are removed, the relative optimal locations should be distributed according to a uniform distribution in $[0,1]$.

In the validation step,  two sets of windows of size $w_v$ are chosen from both markings,  \autoref[see][]{fig:win-comparison}. In the first set, pairs of windows are extracted from the two markings using the optimal vertical shift as determined in the first step, whereas for the second set the windows are extracted using a different (out-of-phase) shift.

@@

Same-shift pairs of length $w_v$ are defined in \citet{hadler} as all 
pairs $(s_i^{(1)}, s_i^{(2)})$ with integer values $i$ for which both $s_i^{(1)}$ and  $s_i^{(2)}$ are defined. Similarly, different-shift pairs are defined as $(s_i^{(1)}, s_{-i-1}^{(2)})$ for all $i$ where both $s_i^{(1)}$ and  $s_{-i-1}^{(2)}$ are defined (see \autoref{sketch-same-diff}).

@@

```{r win-comparison, warning=FALSE, fig.width=8, fig.height=3.5, fig.cap="Two same-source markings. For convenience, the markings are moved into phase on the left and out-of phase on the right. In-phase (left) and out-of-phase (right) samples are shown by the light grey background. The Chumbley-score is based on a Mann-Whitney U test of the correlations derived from these two sets of samples."}

@@

For both same- and different-shift pairs correlations between the  markings are calculated. 
The intuition here is that for two markings from the same-source the correlation for the in-phase sample should be high, while the correlations of the out-of-phase sample  provide a measure for the base-level correlation for non-matching marks of a given length $w_v$. The Chumbley score is then computed as a Mann Whitney U statistic to compare between in-phase sample and out-of-phase sample. 
In the original method proposed in \citet{chumbley} both in-phase and out-of-phase sample are extracted randomly, whereas \citet{hadler} proposed the above specified deterministic rules for both samples to make the resulting score deterministic while simultaneously avoiding overlaps within selected marks to ensure independence.

@@

Looking closer at \autoref{eq.start}, we see that by definition, some number of tests will fail to produce a result. The problem of failed tests is first mentioned in \citet{afte-chumbley}. Unfortunately, the authors do not provide any percentage of how many tests failed for their data.
The algorithm fails to produce for two reasons: either the number of eligible same-shift pairs is zero, or the number of different-shift pairs is zero.

The number of same-shift pairs will be zero, if the optimal locations $t_1^{o}$ and $t_2^{o}$ are so far apart, that no segments of size $w_v$ are left on the same sides of the optimal locations, i.e. $t_1^{o} < w_v$ and $t_2^{o} > T_2-w_o-w_v$ or  $t_1^{o} < T_1- w_o - w_v$ and $t_2^{o} < w_v$, i.e. we have a failure rate of 
\[
P\left( t_1^{o} < w_v \ \cap \ t_2^{o} > T_2-w_o-w_v\right) + P\left( t_1^{o} < T_1- w_o - w_v \ \cap \ t_2^{o} < w_v\right).
\]
While we can assume that once (sub-)class characteristics are removed, optimal locations $t_i^{o}$ are uniformly distributed across the length of the profile, we cannot assume that $t_1^o$ and $t_2^o$ are independent of each other. In particular, for same-source profiles, we would expect a strong dependency between these locations, in which case a large difference between locations is unlikely. However, for different-source matches, we can assume that  locations are independent.
In that case, we expect a test to fail with a probability of $\frac{2 w_v^2}{(T_1-w_o)(T_2-w_o)}$. For an average length of $T_i$ of 1200 pixels, $w_o = 120$ pixels and $w_v = 30$ pixels this probability is about 0.0015.

The number of possible different-shift pairs also depends on the location of  the  optimal locations $t_1^o$ and $t_2^o$. Whenever the optimal locations are close to the boundaries, the number of possible pairings decreases and reaches zero, if $t_i^{(o)} < w_v$ or  $t_i^{(o)} > T_i-w_o- w_v$. Assuming a correlation between optimal locations $t_1^o$ and $t_2^o$ of close to one for same-source profiles, this results in an expected rate of failure of $2 w_v / (T_i-w_o)$, or about 5.6\% for an average length of $T_i$ of 1200 pixels, $w_o = 120$ pixels and $w_v = 30$ pixels. Assuming independence in the optimal locations for different-source profiles the expected probability for a failed test is, again, $\frac{2 w_v^2}{(T_1-w_o)(T_2-w_o)}$.

@@

We then define the pairs for different-shifts as
\begin{equation}\label{eq.diff2}
\left(s_{(j)}^{(1)}, s_{(I-j+1)}^{(1)} \right) \text{ for } j = 
\begin{cases}
1, ..., I & \text{ for even } I \\
1, ..., (I-1)/2, (I-1)/2 + 2, ..., I & \text{ for odd } I
\end{cases},
\end{equation}
i.e.\ for an odd number of same-shift correlations, we skip the middle pair for the different-shift correlations (see \autoref{sketch-diff-2}).
This pairing ensures that the number of different-shift pairings is the same or at most one less than the number of same-shift pairings in all tests.
In the remainder of the paper, we will refer to the algorithm defined by \citet{hadler} as \textbf{(CS1)} and the suggested modified algorithm as \textbf{(CS2)} and compare their performance on the available scans of the Hamby study.
\begin{figure}[hbtp]
\centering
\includegraphics[width=.7\textwidth]{images/sketch-diff-2.png}
\caption{\label{sketch-diff-2}Sketch of adjusted different-shift pairings. At most one of the same-shift pairings can not be matched with a different-shift pair under algorithm (CS2). }
\end{figure}

@@

@@

## Setup
Both algorithms (CS1) and (CS2) are implemented in R \citep{R}. (CS1) is available from package `toolmaRk` \citep{toolmark}, (CS2) is available from a modified version of the `toolmaRk` package available from GitHub (\url{https://github.com/heike/toolmaRk}). We applied both methods to all pairwise land-to-land comparisons of the Hamby scans provided by NIST for a total of 85,491 land-to-land comparisons.

@@

Initially, the default settings suggested in \citet{hadler} were used: $w_o = 120$ pixels or about 190 $\mu m$ (ten percent of the average length of profiles) and coarseness $c = 0.25$. 
\autoref{fig:fails} shows the percentage of failed tests among the 85,491 land-to-land comparisons of the NIST data for different values of the validation window size $w_v$. For same-source lands up to 12.5 percent of the tests fail using CS1. The highest percentage of failed tests under CS2 is 1.3% for different-source tests using a validation window size $w_v$ of 60 pixels.  \hh{Rates of expected failures are based on simulation runs using covariances between locations of same-source profiles of 0.854, and 0.120 for locations from different-source profiles, matching observed covariances for the Hamby scans. }\hh{ Observed failure rates are higher than expected. This might be due to remaining sub-class structure at a coarseness of 0.3 resulting in a distribution of optimal locations different from the assumed uniform. }

@@

\autoref{fig:profile-sketch} gives an overview of the effect of different coarseness parameters: from left to right, coarseness levels $c$ are varied in steps of 0.05 from 0.1 to 0.3. The top row shows  resulting signatures after smoothing the profile shown in Figure~\ref{fig:sigs-profiles} with different levels of coarseness. 
The histograms in the bottom row show the relative optimal location $t^*$. Optimal locations are distributed uniformly once (sub-)class characteristics are removed. However, for coarseness values of $c > 0.20$ we see quite distinct boundary effects: optimal locations $t^*$ are found at the very extreme ends of a profile more often than one would expect based on a uniform distribution.

@@

<!-- The results suggest that the Nominal type I error $\alpha$ value shows dependence on the size of the window of optimization. For a given window of optimization the actual Type I error is comparable to the nominal level for only a select few validation window sizes and for comparable validation window sizes of 30 and 50 as done here, the actual type I error does not seem to vary as much as it varies with the optimization window sizes .
A Test Fail, i.e. tests in which a particular parameter setting did not return a valid result, happens, when the shift to align two signatures is so large, that the remaining overlap is too small to accommodate  windows for validation, depends on whether known-match or known non-matches has predictive value, with test results from different-sources having a much higher chance to fail. On conducting an analysis of all known bullet lands using the adjusted chumbley algorithm, Type II error was identified to be least bad for window of validation 30 and window of optimization 120. In case of unsmoothed raw marks (profiles), Type II error increases with the amount of smoothing and least for LOWESS smoothing coarseness value about 0.25 or 0.3. In an effort to identify the level of adaptiveness of the algorithm, comparisons were made between signatures and profiles. Their comparison with respect to validation window size for a fixed optimization window size suggested that, profiles have a total error (i.e all incorrect classification of known-matches and known non-matches) greater than or equal to the total error of signatures for all sizes of validation window. Profiles also fail more number of times than signatures in a test fail (for different coarseness keeping windows fixed and also for different validation windows keeping coarseness fixed) which lets us conclude that the behaviour of the algorithm for the profiles instead of pre-processed signatures is not better. Finally it should be noted that the current version of the adjusted chumbley algorithm seems to fall short when compared to other machine-learning based methods \citet{aoas}, and some level of modification to the deterministic algorithm needs to be identified and tested that would reduce the number of incorrect classifications.

@@

# Appendix
\begin{appendix}

\section{Scenarios of failed tests}

Figures \ref{same-shift-failure} and \ref{diff-shift-failure} show scenarios in which the deterministic Chumbley score can fail. 
In \autoref{same-shift-failure} both CS1 and CS2 fail, because the lag between optimal locations is so large that no same-shift pairs can be found once the signatures are aligned. These failures are inherent to the Chumbley Score and cannot be prevented.
\begin{figure}[hbtp]
\centering
\includegraphics[width=0.62\textwidth]{images/same-shift-failure.png}

\caption{\label{same-shift-failure}Sketch of same-shift pairings. When the lag between optimal reference points is too large to accommodate a validation window in noth signatures, both CS1 and CS2 fail.}
\end{figure}

\autoref{diff-shift-failure} shows two situations, in which CS1 fails to identify any different-shift pairings. This happens, when both relative locations are close to either end of the signature. CS2 can still be computed in this situation as long as there are at least two same-shift pairs identified.

\caption{\label{diff-shift-failure}Sketch of different-shift pairings. For CS1 no different-shift pairs can be identified, resulting in a failed test.}
\end{figure}

<!--
\begin{figure}[hbtp]
\centering
\includegraphics[width=.7\textwidth]{images/CS1-failure-CS2-success.png}
\caption{\label{CS1-failure-CS2-success}Sketch of a case where CS1 fails but CS2 does not fail.}
\end{figure}
-->

\end{appendix}
